Semi-permeable vesicles: mathematical formulation, numerics and physics

### 2020-03-13 vesicle_code/docs/permeableVes/main.tex
picking at paper

\section{Problem formulation} \label{sc:formulate}
Consider a two-dimensional semi-permeable vesicle suspended in an
unbounded viscous fluid domain, subjected to an imposed flow
$\uu_\infty(\xx)$, for any $\xx \in \RR^2$.  Assume that the interior
and exterior fluids have the same viscosity $\mu$. Let $\xx$ be the
position of the interface $\gamma$, $\uu$ be the fluid velocity, and
$p$ be the pressure. In the vanishing Reynolds number limit, the
governing equations for the ambient fluid are
%
\begin{subequations}
\begin{alignat}{3}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu &= 0, 
  &&\xx \in \RR^2 \setminus \gamma, \\
% 
  \nabla \cdot \uu &= 0,  &&\xx \in \RR^2 \setminus \gamma, \\
%
  \uu(\xx) &\rightarrow \uu_\infty(\xx),
    \qquad &&\norm{\xx} \rightarrow \infty.
\end{alignat}
\end{subequations}%
%
Letting $\nn$ be the outward normal to $\gamma$, $\jump{\cdot}$ denote
the jump across the interface, and $T$ denote the hydrodynamic stress
tensor, the boundary conditions along the vesicle are
\begin{align}
  \label{eqn:jump1}
  \jump{T\cdot\nn} &= \ff_\text{mem}, \\
  \label{eqn:jump2}
  \jump{\uu} &= 0,
\end{align}
where $\ff_\text{mem}$ is the standard merrane force that includes a
bending and a tension term. In particular, $\ff_\text{mem} =
\ff_\text{ben} + \ff_\text{ten}$, where
\begin{align}
  \label{eqn:merraneForces}
  \ff_\text{ben} = -\kappa_b \xx_{ssss}, \qquad
  \ff_\text{ten} = (\sigma \xx_s)_s,
\end{align}
where $\kappa_b$ is the bending stiffness and $\sigma$ is a Lagrange
multiplier, which acts like the tension, which is used to satisfy the
inextensibility condition.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle merrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\xx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds .
\end{align}
The time derivative of $A(t)$ includes terms due to the time derivative
of $\xx \cdot \nn$, and a term due to interface
stretching~\cite{lai-tse-hua2008}. However, this second term is zero
because of the local inextensibility condition. Therefore, 
\begin{align}
  \dot{A}(t) &= \frac{1}{2} \int_{\gamma} \pderiv{}{t} 
    (\xx \cdot \nn) \, ds = 
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
The second integrand can be computed as
\begin{align}
  \int_\gamma (\xx \cdot \dot{\nn})\, ds 
    = -\int_\gamma \left(\xx^\perp \cdot \dot{\xx}_s\right)\, ds 
    = \int_\gamma \left(\xx_s^\perp \cdot \dot{\xx}\right)\, ds 
    = \int_\gamma (\dot{\xx} \cdot \nn)\, ds.
\end{align}
Therefore,
\begin{align}
  \dot{A}(t) &= \int_\gamma (\dot{\xx} \cdot \nn)\, ds 
  = \int_\gamma \left(\beta (\ff \cdot \nn)\nn + \SS[\ff]\right) 
      \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where the last equality comes follows from the divergence theorem.
Finally, using integration by parts and the Frenet-Serret identities, we
have
\begin{align*}
  \dot{A}(t) &= \beta \int_\gamma \left( -\kappa_b \xx_{ssss} + 
    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
  &= \beta \int_\gamma \left(-\kappa_b \xx_{sss} + \sigma \xx_s 
    \right) \cdot \kappa \xx_s \, ds \\
  &= \beta \int_\gamma \left(\kappa_b \xx_{ss} \cdot 
    (\kappa \xx_s)_s + \kappa \sigma \right) \, ds \\
  &= \beta \int_\gamma \left(\kappa_b \kappa (\xx_{ss} \cdot \xx_{ss}) + 
    \kappa_b \kappa_s (\xx_{ss} \cdot \xx_s) + \kappa \sigma \right) 
    \, ds \\
  &= \beta \int_\gamma \left(\kappa_b \kappa^3 + \kappa \sigma \right) 
    \, ds.
\end{align*}

@@

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:merraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\xx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\xx]\ff &= -\frac{d^4}{ds^4} \ff, \\
  T[\xx]\sigma &= (\sigma \xx_s)_s, \\
  D[\xx]\ff &= \xx_s \cdot \ff_s, \\
  P[\xx]\ff &= (\ff \cdot \nn) \nn. 
\end{align}
Here, the normal vector $\nn$ is the outward unit normal of the
$\gamma$, and $d/ds$ is its arclength derivative. To simplify notation,
we write $B\ff$ to mean $B[\xx]\ff$, and use similar notation for the
other operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\xx}{dt} &= \uu_{\infty}(\xx) + 
  \beta P(B\xx + T\sigma) + \SS(B\xx + T\sigma), \\
  D \frac{d\xx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t_N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\xx^{N+1} - \xx^N}{\Delta t} = \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\xx^{N+1} = 1.
\end{align}

@@

@@

At steady state,
\begin{align}
    \dot{A}(t) &= \beta \int_{\gamma}\left(\ff\cdot\nn\right)ds =0.
\end{align}
We now replace $\ff$ with its definition and apply integration by parts
\begin{align}
   0=\beta\int_{\gamma} 
    (-\kappa_b\xx_{ssss}+(\sigma\xx_s)_s)\cdot\nn\, ds =
   \beta\int_{\gamma}\left(\kappa_b\xx_{sss}-\sigma\xx_s\right)\cdot
     \nn_s\, ds
\end{align}
From the Frenet-Serret formulas, we have 
\begin{align}
     -\beta\int_{\gamma}
     \left(\kappa_b\xx_{sss}-\sigma\xx_s\right)\cdot\kappa\xx_s ds &=0.
\end{align}

Applying integration by parts and Frenet-Serret again
\begin{align}
     \beta\int_{\gamma}
     \kappa_b\kappa\left(\xx_{ss}\cdot\xx_{ss}\right)ds + \beta
     \int_{\gamma} \kappa_b \kappa_s(\xx_{ss} \cdot \xx_s)ds +\beta\int_{\gamma}\kappa\sigma ds  &= 0,\\
    \beta\int_{\gamma} \kappa_b\kappa(\kappa\nn \cdot
     \kappa\nn)+ \beta \int_{\gamma} \kappa_b \kappa_s(\xx_{ss} \cdot
     \xx_s)ds +\beta\int_{\gamma}\kappa\sigma ds &= 0,
\end{align}
which we re-write as
\begin{align}
   \beta\int_{\gamma}\kappa_b\kappa^3 ds + \beta \int_{\gamma} \kappa_b
   \kappa_s(\xx_{ss} \cdot \xx_s)ds +\beta\int_{\gamma} \kappa\sigma ds = 0.
\end{align}
Applying Frenet-serret again,
\begin{align}
    \beta\int_{\gamma}\kappa_b\kappa^3 ds + \beta \int_{\gamma} \kappa_b
    \kappa_s(\kappa\nn \cdot \TT)ds +\beta\int_{\gamma} \kappa\sigma ds = 0, \\
    \beta\int_{\gamma}\kappa_b\kappa^3 ds +\beta\int_{\gamma} \kappa\sigma ds  = 0.
\end{align}

>>>>

\section{Problem formulation} \label{sc:formulate}
Consider a two-dimensional semi-permeable vesicle suspended in an
unbounded viscous fluid domain, subjected to an imposed flow
$\uu_\infty(\xx)$, for any $\xx \in \RR^2$.  Assume that the interior
and exterior fluids have the same viscosity $\mu$. Let $\xx$ be the
position of the interface $\gamma$, $\uu$ be the fluid velocity, and
$p$ be the pressure. In the vanishing Reynolds number limit, the
governing equations for the ambient fluid are
%
\begin{subequations}
\begin{alignat}{3}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu &= 0, 
  &&\xx \in \RR^2 \setminus \gamma, \\
% 
  \nabla \cdot \uu &= 0,  &&\xx \in \RR^2 \setminus \gamma, \\
%
  \uu(\xx) &\rightarrow \uu_\infty(\xx),
    \qquad &&\norm{\xx} \rightarrow \infty.
\end{alignat}
\end{subequations}%
%
Letting $\nn$ be the outward normal to $\gamma$, $\jump{\cdot}$ denote
the jump across the interface, and $T$ denote the hydrodynamic stress
tensor, the boundary conditions along the vesicle are
\begin{align}
  \label{eqn:jump1}
  \jump{T\cdot\nn} &= \ff_\text{mem}, \\
  \label{eqn:jump2}
  \jump{\uu} &= 0,
\end{align}
where $\ff_\text{mem}$ is the standard merrane force that includes a
bending and a tension term. In particular, $\ff_\text{mem} =
\ff_\text{ben} + \ff_\text{ten}$, where
\begin{align}
  \label{eqn:merraneForces}
  \ff_\text{ben} = -\xx_{ssss}, \qquad
  \ff_\text{ten} = (\sigma \xx_s)_s.
\end{align}
Since the bending stiffness only sets the time scale, we let $k_b = 1$
throughout this paper. The Lagrange multiplier $\sigma$ is used to
satisfy the inextensibility condition.  Finally, the tension of the
vesicle is
\begin{align}
  \Lambda = \sigma + 1.5 \kappa^2,
\end{align}
where $\kappa$ is the membrane curvature.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle membrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\xx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds .
\end{align}
The time derivative of $A(t)$ includes terms due to the time derivative
of $\xx \cdot \nn$, and a term due to interface
stretching~\cite{lai-tse-hua2008}. However, this second term is zero
because of the local inextensibility condition. Therefore, 
\begin{align}
  \dot{A}(t) &= \frac{1}{2} \int_{\gamma} \pderiv{}{t} 
    (\xx \cdot \nn) \, ds = 
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
The second integrand can be computed as
\begin{align}
  \int_\gamma (\xx \cdot \dot{\nn})\, ds 
    = -\int_\gamma \left(\xx^\perp \cdot \dot{\xx}_s\right)\, ds 
    = \int_\gamma \left(\xx_s^\perp \cdot \dot{\xx}\right)\, ds 
    = \int_\gamma (\dot{\xx} \cdot \nn)\, ds.
\end{align}
Therefore,
\begin{align}
  \dot{A}(t) &= \int_\gamma (\dot{\xx} \cdot \nn)\, ds 
  = \int_\gamma \left(\beta (\ff \cdot \nn)\nn + \SS[\ff]\right) 
      \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where the last equality comes follows from the divergence theorem.
Finally, using integration by parts and the Frenet-Serret identities, we
have
\begin{align}
  \dot{A}(t) &= \beta \int_\gamma \left( -\xx_{ssss} + 
    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
  &= \beta \int_\gamma \left(-\xx_{sss} + \sigma \xx_s 
    \right) \cdot \kappa \xx_s \, ds \\
  &= \beta \int_\gamma \left(\xx_{ss} \cdot 
    (\kappa \xx_s)_s + \kappa \sigma \right) \, ds \\
  &= \beta \int_\gamma \left(\kappa (\xx_{ss} \cdot \xx_{ss}) + 
    \kappa_s (\xx_{ss} \cdot \xx_s) + \kappa \sigma \right) 
    \, ds \\
  &= \beta \int_\gamma \left(\kappa^3 + \kappa \sigma \right) 
    \, ds.
\end{align}
Finally, using the relationship between the Lagrange multiplier
$\sigma$, and the tension $\Lambda$, the area of a semi-permeable
vesicle is governed by
\begin{align}
  \dot{A}(t) = \beta \int_\gamma \left(
    -\frac{\kappa^3}{2} + \kappa \Lambda \right) \, ds.
\end{align}

@@

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:merraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\xx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\xx]\ff &= -\frac{d^4}{ds^4} \ff, \\
  T[\xx]\sigma &= (\sigma \xx_s)_s, \\
  D[\xx]\ff &= \xx_s \cdot \ff_s, \\
  P[\xx]\ff &= (\ff \cdot \nn) \nn. 
\end{align}
Here, the normal vector $\nn$ is the outward unit normal of the
$\gamma$, and $d/ds$ is its arclength derivative. To simplify notation,
we write $B\ff$ to mean $B[\xx]\ff$, and use similar notation for the
other operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\xx}{dt} &= \uu_{\infty}(\xx) + 
  \beta P(B\xx + T\sigma) + \SS(B\xx + T\sigma), \\
  D \frac{d\xx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t^N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\xx^{N+1} - \xx^N}{\Delta t} = \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\xx^{N+1} = 1.
\end{align}

@@

We consider a single semi-permeable vesicle in a quiescent flow. For
impermeable vesicles, the vesicle relaxes to a shape with constant
negative tension~\cite{kra-win-sei-lip1996}.

@@

At steady state,
\begin{align}
    \dot{A}(t) &= \beta \int_{\gamma}\left(\ff\cdot\nn\right)ds =0.
\end{align}
We now replace $\ff$ with its definition and apply integration by parts
\begin{align}
   0=\beta\int_{\gamma} 
    (-k_b\xx_{ssss}+(\sigma\xx_s)_s)\cdot\nn\, ds =
   \beta\int_{\gamma}\left(k_b\xx_{sss}-\sigma\xx_s\right)\cdot
     \nn_s\, ds
\end{align}
From the Frenet-Serret formulas, we have 
\begin{align}
     -\beta\int_{\gamma}
     \left(k_b\xx_{sss}-\sigma\xx_s\right)\cdot\kappa\xx_s ds &=0.
\end{align}

Applying integration by parts and Frenet-Serret again
\begin{align}
     \beta\int_{\gamma}
     k_b\kappa\left(\xx_{ss}\cdot\xx_{ss}\right)ds + \beta
     \int_{\gamma} k_b \kappa_s(\xx_{ss} \cdot \xx_s)ds +\beta\int_{\gamma}\kappa\sigma ds  &= 0,\\
    \beta\int_{\gamma} k_b\kappa(\kappa\nn \cdot
     \kappa\nn)+ \beta \int_{\gamma} k_b \kappa_s(\xx_{ss} \cdot
     \xx_s)ds +\beta\int_{\gamma}\kappa\sigma ds &= 0,
\end{align}
which we re-write as
\begin{align}
   \beta\int_{\gamma}k_b\kappa^3 ds + \beta \int_{\gamma} k_b
   \kappa_s(\xx_{ss} \cdot \xx_s)ds +\beta\int_{\gamma} \kappa\sigma ds = 0.
\end{align}
Applying Frenet-serret again,
\begin{align}
    \beta\int_{\gamma}k_b\kappa^3 ds + \beta \int_{\gamma} k_b
    \kappa_s(\kappa\nn \cdot \TT)ds +\beta\int_{\gamma} \kappa\sigma ds = 0, \\
    \beta\int_{\gamma}k_b\kappa^3 ds +\beta\int_{\gamma} \kappa\sigma ds  = 0.
\end{align}
